Safe Transferable Regions

### 2016-07-02 paper/type-system.tex
added op semantics
added appendix and moved some material there
removed constructor

The purpose of \name's region type system is to enforce the key
invariant required for memory safety, namely that an object $o_1$ in a
region $R_1$ contains a reference to an object $o_2$ in $R_2$, only if
$R_2$ is guaranteed to outlive $R_1$. 
% Intuitively, the invariant can
% be enforced by (a). tracking \emph{outlives} relationships between
% various regions in the program (b). tagging the C\# type of every
% object in the program with its allocation region, and (c).  ensuring
% that, when a reference is created, its target object ($o_2$) is
% allocated in a region that is known to be in outlives relationship
% with the object containing the reference. 
We now formally develop this intuition via \fbname (\FB), our
explicitly typed core language (with region types) that incorporates
the features introduced in the previous section. \fbname builds on the
Featherweight Generic Java (FGJ)~\cite{fgj} formalism, and reuses
notations and various definitions from~\cite{fgj}, such as the
definition of type well-formedness for the core (region-free)
language.

@@

Fig~\ref{fig:fb-syntax} describes the syntax of \FB.
We refer to the class types of FGJ as \emph{core types}.
%
The following definition of \C{Pair} class in \FB illustrates some
of the key elements of the formal language\footnote{The symbol
$\extends$ should be read \emph{extends}, and the symbol $\outlives$
stands for \emph{outlives}}:
\begin{codejava}[mathescape=true]
class Pair<a $\extends$ Object, b $\extends$ Object>
          <$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a \conj \rho_2 \outlives \rho^a$> $\extends$ Object<$\rho^a$> {
  a@$\rho_1$ fst; 
  b@$\rho_2$ snd;
  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
    super(); this.fst = fst; this.snd = snd;
  }
  a@$\rho_1$ getFst() { 
    return this.fst; 
  }
}
\end{codejava}
\vspace*{-0.15in}

@@

Well-formedness and typing rules of \fbname establish the conditions
under which a region type is considered well-formed, and an expression
is considered to have a certain region type, respectively.
Fig.~\ref{fig:fb-staticsem} contains an illustrative subset of such
rules\footnote{The full formal development can be found in the
supplement}. The rules refer to a context ($\A$), which is a tuple of:
\begin{itemize}
\item A set ($\rhoenv \in 2^\rgn$) of static identifiers of regions
that are estimated to be live,
\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
variables to their bounds\footnote{A bound of a type variable
($\tyvar$) in FGJ~\cite{fgj} is the class ($\fgjN$) that type variable
was declared to extend.}, and
\item A constraint formula ($\phicx$) that captures the outlives
constraints on regions in $\rhoenv$.
\end{itemize}
$\A$ also contains $\rhomap$, 
% a map ($\rhomap \in \rgn \mapsto
% \{\CLOSED,\OPEN,\XFERRED\}$) from unique region identifiers of
% transferable regions to their current state in the FSM from
% Fig.~\ref{fig:region-fsm}.
which is primarily an artifact to facilitate the type safety proof,
and can be ignored while type checking user-written programs
in \FB. The context for the expression typing judgment also includes:
\begin{itemize}
\item A type environment ($\env \in x \mapsto \tau$) that contains the
type bindings for variables in scope, and 
\item The static identifier ($\rgn^a$) of the allocation context for
the expression is being typechecked.  
\end{itemize}
Like the judgments in FGJ~\cite{fgj}, all the judgments defined by the
rules in Fig.~\ref{fig:fb-staticsem} are implicitly parameterized on a
class table ($CT \in cn \mapsto D$) that maps class names to their
definitions in \FB.
% \begin{figure}

% \begin{codejava}
% class Region<a $\extends$ Object><${\rhoalloc}$> $\extends$ Object<$\rhoalloc$>{
%   <${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk;
%   Region(<${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk) {
%     super(); 
%     this.thunk = thunk;
%   }
%   unit transfer$\inang{\rho}$(unit u) {
%     return u;
%   }
% }
% \end{codejava}

% \caption{The stub used for $\RgnZ$ class}
% \label{fig:region-stub}
% \end{figure}

@@

The supplement defines a small-step operational semantics for \fbname
via a five-place reduction relation:
% \vspace*{-0.15in}
\begin{smathpar}
  \redstoo{(e,\rhomap)}{(e',\rhomap')}
\end{smathpar}
The reduction judgment should be read as following: given a set
($\rhoenv$) of regions that are currently live, and a map ($\rhomap$)
from unique identifiers of transferable regions to their current
states (Fig.~\ref{fig:region-fsm}), the expression $e$ reduces to
$e'$, while updating $\rhomap$ to $\rhomap'$. The semantics gets
``stuck'' if $e$ attempts to access an object whose allocation region
is not present in $\rhoenv$, or if $e$ tries to \C{open} a
transferable region, whose identifier is not mapped to a state by
$\rhomap$.  On the other hand, if $e$ attempts to commit an operation
on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$).

>>>>

The purpose of \name's region type system is to enforce the key
invariant required for memory safety, namely that an object $o_1$ in a
region $R_1$ contains a reference to an object $o_2$ in $R_2$, only if
$R_2$ is guaranteed to outlive $R_1$. 
We now formally develop this intuition via \fbname (\FB), our
explicitly typed core language (with region types) that incorporates
the features introduced in the previous section. \fbname builds on the
Featherweight Generic Java (FGJ)~\cite{fgj} formalism, and reuses
notations and various definitions from~\cite{fgj}, such as the
definition of type well-formedness for the core (region-free)
language.

@@

Fig~\ref{fig:fb-syntax} describes the syntax of \FB.
We refer to the class types of FGJ as \emph{core types}.
%
The following definition of \C{Pair} class in \FB illustrates some
of the key elements of the formal language\footnote{The symbol
$\extends$ should be read \emph{extends}, and the symbol $\outlives$
stands for \emph{outlives}}:
\begin{codejava}[mathescape=true]
class Pair<a $\extends$ Object, b $\extends$ Object>
          <$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a \conj \rho_2 \outlives \rho^a$> $\extends$ Object<$\rho^a$> {
  a@$\rho_1$ fst; 
  b@$\rho_2$ snd;
  a@$\rho_1$ getFst() { 
    return this.fst; 
  }
}
\end{codejava}
% \vspace*{-0.15in}

%  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
%    super(); this.fst = fst; this.snd = snd;
%  }

@@

Well-formedness and typing rules of \fbname establish the conditions
under which a region type is considered well-formed, and an expression
is considered to have a certain region type, respectively.
Fig.~\ref{fig:fb-staticsem} contains an illustrative subset of such
rules\footnote{The full formal development can be found in the
supplement}. The rules refer to a context ($\A$), which is a tuple of:
\begin{itemize}
\item A set ($\rhoenv \in 2^\rgn$) of static identifiers of regions
that are estimated to be live,
\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
variables to their bounds\footnote{A bound of a type variable
($\tyvar$) in FGJ~\cite{fgj} is the class ($\fgjN$) that type variable
was declared to extend.}, and
\item A constraint formula ($\phicx$) that captures the outlives
constraints on regions in $\rhoenv$.
\end{itemize}
$\A$ also contains $\rhomap$, 
% a map ($\rhomap \in \rgn \mapsto
% \{\CLOSED,\OPEN,\XFERRED\}$) from unique region identifiers of
% transferable regions to their current state in the FSM from
% Fig.~\ref{fig:region-fsm}.
which is primarily an artifact to facilitate the type safety proof,
and can be ignored while type checking user-written programs
in \FB. The context for the expression typing judgment also includes:
\begin{itemize}
\item A type environment ($\env \in x \mapsto \tau$) that contains the
type bindings for variables in scope, and 
\item The static identifier ($\rgn^a$) of the allocation context for
the expression is being typechecked.  
\end{itemize}
Like the judgments in FGJ~\cite{fgj}, all the judgments defined by the
rules in Fig.~\ref{fig:fb-staticsem} are implicitly parameterized on a
class table ($CT \in cn \mapsto D$) that maps class names to their
definitions in \FB.

@@

Fig.~\ref{fig:fb-opsem} in the appendix defines a small-step operational semantics for \fbname
via a five-place reduction relation:
% \vspace*{-0.15in}
\begin{smathpar}
  \redstoo{(e,\rhomap)}{(e',\rhomap')}
\end{smathpar}
The reduction judgment should be read as following: given a set
($\rhoenv$) of regions that are currently live, and a map ($\rhomap$)
from unique identifiers of transferable regions to their current
states (Fig.~\ref{fig:region-fsm}), the expression $e$ reduces to
$e'$, while updating $\rhomap$ to $\rhomap'$. The semantics gets
``stuck'' if $e$ attempts to access an object whose allocation region
is not present in $\rhoenv$, or if $e$ tries to \C{open} a
transferable region, whose identifier is not mapped to a state by
$\rhomap$.  On the other hand, if $e$ attempts to commit an operation
on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$).
